On the surprising effectiveness of a simple matrix exponential derivative approximation, with application to global SARS-CoV-2

### 2023-09-13 manuscript/revision_letter.tex

\begin{enumerate}
	\item The authors spend 3 out of 7 pages in the main text and three corresponding sections in the Appendix discussing the refined error and the proof, while postpone the application with the affine-corrected error to future work. I would expect more than a relatively abstract error comparison figure for this issue. For example, a simple numerical simulation similar to the last paragraph in Application to compute the posterior likelihood ratio or posterior density ratio of the corrected and the uncorrected methods, because the authors eventually applied the method within a Bayesian framework.
\end{enumerate}

@@

\begin{enumerate}	
	  \setcounter{enumi}{1}
	\item I am curious about the selection of the extra two additional Chinese provinces and 11 additional countries when fitting the mixed-effects model and how the inclusion of the extra regions improve the fitting. Is there any criteria to determine the inclusion?
\end{enumerate}

@@

\begin{enumerate}	
\setcounter{enumi}{2}
	\item For the mixed-effects model, do the authors assume independent predictors for the fixed effects? It seems to me that, in Fig. 2, the air traffic proximity and the intracontinental proximity are highly correlated, which can be the reason why $\theta_2$ is not statistically significant. Then what’s the justification for the predictor selection?
\end{enumerate}

@@

\begin{enumerate}	
\setcounter{enumi}{3}	
	\item Am I correct that the total number of particles (i.e., samples) in the MCMC framework is 2,000? The ESS for $\tau$ is extremely small, do the authors have any explanations?
\end{enumerate}

@@

\begin{enumerate}	
\setcounter{enumi}{4}	
	\item What does the random effect matrix look like? The authors indicate that air traffic proximity is the only significant fixed-effect predict, then I would wonder how much does the random effect contribute to the model?
\end{enumerate}

\setcounter{figure}{4}
\begin{figure}[t]
	\centering
	\includegraphics[width=0.8\linewidth]{figures/randEffects.pdf}
	\caption{Posterior means for exponentiated random effects convey expected multiplicative deviations from the portion of the rate attributable to fixed effects for each corresponding element of the generator matrix.  Notably, we infer a roughly 1.81-fold posterior mean increase in the rate of transitions from the US to Hubei, CN, beyond that portion of the rate which may be explained by fixed effects.  Less pronounced are posterior mean multiplicative increases of 1.34 (US to Italy), 1.27 (US to UK), 1.44 (Netherlands to Italy) and 1.21 (Guangdong, CN, to Hubei, CN).}%\label{fig:randEffs}
\end{figure}

@@

\begin{enumerate}	
\setcounter{enumi}{5}	
	\item The ranges for predictor matrices and posterior mean rate matrix are different. However, the color palettes are identical, which makes it extremely hard to compare the impact of each predictor. For example, I cannot see how the sentence ``... but one may also see the influence of the Hubei asymmetry in, e.g., the squares corresponding to travel between Guangdong and Hubei provinces.'' makes sense visually from Fig. 2 and 3. It also doesn’t make sense to me conceptually, because air traffic proximity is the only significant predictor.
\end{enumerate}

@@

\begin{enumerate}	
\setcounter{enumi}{6}	
	\item 
	The color palette in Fig. 4 makes it really hard to distinguish Hubei, other provinces in China, and other countries. It will be great if the authors can highlight Hubei in one color, other provinces in another color with gradients, and other countries in a third color with gradients. And why there are only 29 regions?
\end{enumerate}

@@

\begin{enumerate}	
\setcounter{enumi}{7}	
	\item 
	The proposed method is able to include higher-dimension rate matrix compared with \citet{lemey2009bayesian,lemey2014unifying}. Can the authors then have an external comparison to the results in \citet{lemey2020accommodating} just visually to see whether the color pattern in the phylogenetic tree is matched?
\end{enumerate}

>>>>

1. The authors spend 3 out of 7 pages in the main text and three corresponding sections in the Appendix discussing the refined error and the proof, while postpone the application with the affine-corrected error to future work. I would expect more than a relatively abstract error comparison figure for this issue. For example, a simple numerical simulation similar to the last paragraph in Application to compute the posterior likelihood ratio or posterior density ratio of the corrected and the uncorrected methods, because the authors eventually applied the method within a Bayesian framework.

@@

2. I am curious about the selection of the extra two additional Chinese provinces and 11 additional countries when fitting the mixed-effects model and how the inclusion of the extra regions improve the fitting. Is there any criteria to determine the inclusion?

@@

3. For the mixed-effects model, do the authors assume independent predictors for the fixed effects? It seems to me that, in Fig. 2, the air traffic proximity and the intracontinental proximity are highly correlated, which can be the reason why $\theta_2$ is not statistically significant. Then what’s the justification for the predictor selection?

@@

4. Am I correct that the total number of particles (i.e., samples) in the MCMC framework is 2,000? The ESS for $\tau$ is extremely small, do the authors have any explanations?

@@

5. What does the random effect matrix look like? The authors indicate that air traffic proximity is the only significant fixed-effect predict, then I would wonder how much does the random effect contribute to the model?

\setcounter{figure}{4}
\begin{figure}[t]
	\centering
	\includegraphics[width=0.7\linewidth]{figures/randEffects.pdf}
	\caption{Posterior means for exponentiated random effects convey expected multiplicative deviations from the portion of the rate attributable to fixed effects for each corresponding element of the generator matrix.  Notably, we infer a roughly 1.81-fold posterior mean increase in the rate of transitions from the US to Hubei, CN, beyond that portion of the rate which may be explained by fixed effects.  Less pronounced are posterior mean multiplicative increases of 1.34 (US to Italy), 1.27 (US to UK), 1.44 (Netherlands to Italy) and 1.21 (Guangdong, CN, to Hubei, CN).}%\label{fig:randEffs}
\end{figure}

@@

6. The ranges for predictor matrices and posterior mean rate matrix are different. However, the color palettes are identical, which makes it extremely hard to compare the impact of each predictor. For example, I cannot see how the sentence ``... but one may also see the influence of the Hubei asymmetry in, e.g., the squares corresponding to travel between Guangdong and Hubei provinces.'' makes sense visually from Fig. 2 and 3. It also doesn’t make sense to me conceptually, because air traffic proximity is the only significant predictor.

@@

7.	The color palette in Fig. 4 makes it really hard to distinguish Hubei, other provinces in China, and other countries. It will be great if the authors can highlight Hubei in one color, other provinces in another color with gradients, and other countries in a third color with gradients. And why there are only 29 regions?

@@

8.	The proposed method is able to include higher-dimension rate matrix compared with \citet{lemey2009bayesian,lemey2014unifying}. Can the authors then have an external comparison to the results in \citet{lemey2020accommodating} just visually to see whether the color pattern in the phylogenetic tree is matched?

\begin{reply}
	Thank you for the suggestion.  We now notice a strong correspondence between our updated Figure \ref{fig:tree} and Fig1 of \citet{lemey2020accommodating}.  We have accordingly added the following text to the manuscript:
	\begin{quote}
		After collapsing regions into 8 major blocks, Figure \ref{fig:tree} projects the empirical posterior predictive mode of these blocks onto the phylogenetic tree $\phylogeny$.  The general pattern looks similar to that of Figure 1 from \cite{lemey2020accommodating}, although the geographic blocking scheme differs slightly.
	\end{quote}
\end{reply}

\begin{reply}
	Again, we wish to thank you for taking the time to review our manuscript.  We are confident that the revised manuscript is much improved as a result of your efforts.
\end{reply}
